A Lyapunov Analysis of Momentum Methods \\ in Optimization

### 2016-10-28 LyapDraft.tex
Update LyapDraft.tex

\section{The dynamical view of momentum methods}

@@

\section{Rate-generating Lyapunov Functions}
Lyapunov's method~\cite{Lyapunov} is based on the idea of constructing a positive definite quantity $\E: \X \rightarrow \R$ which decreases along the trajectories of the dynamical system $\dot X_t = v(X_t)$:
\begin{equation*}
\dot \E(X_t) =\langle \nabla \E(X_t), v(X_t)\rangle < 0.
\end{equation*}
The existence of such a \emph{Lyapunov function} guarantees that the dynamical system converges: if the function is positive yet strictly decreasing along all trajectories, then the dynamical system must eventually approach a region where $\E(X)$ is minimal.  If this region coincides with the stationary points of the dynamics, then all trajectories must converge to a stationary point.

@@

\subsection{The Quasi-Monotone Subgradient Method}
\label{Sec:QuasiMono}
The quasi-monotone subgradient method was developed by Nesterov~\cite{Nesterov15} as an alternative to the dual averaging method -- though both achieve the $O(1/\sqrt{k})$ lower bound for the class of methods with bounded subgradients. The advantage of the quasi-monotone method however is that a convergence rate guarantee can be shown for the entire sequence of iterates instead of the average (or minimum) iterate, as is the case with the dual averaging method in the non-Euclidean setting. {\em We begin by studying the quasi-monotone method in the situation where we have full gradients, which we assume to be bounded}. We also set the dual averaging term $\gamma_k \equiv 1$ for convenience and give an analysis of the full algorithm in Appendix~\ref{App:ProofQuasiGen}.  With these minor modifications, the algorithm can be understood as the explicit-Euler method applied to~\eqref{Eq:Z} and the implicit-Euler method applied to~\eqref{Eq:X} (where now, we take $e^{\beta_t} = A_{k+1}$):

@@

\subsection{Other Discretizations}
We give examples of two other algorithms that are the result of ``discretizing'' the momentum dynamic~\eqref{Eq:DynaBeta}, and analyze them using the discretized Lyapunov function~\eqref{Eq:DiscLyapFunc1}. The proofs of these results can be found in the Appendix~\ref{App:OtherDiscret}. The first method is the result of applying explicit-Euler method to~\eqref{Eq:X} and the implicit-Euler method to~\eqref{Eq:Z}:
\begin{algorithm}[H]
\caption{Method 1}
{\bf Assumptions:} $f$ is smooth and $\X$ is convex and compact. \\
Let $A_0 = 1$, $x_0 = z_0$, $\alpha_k = A_{k+1} - A_k$ and $\tau_k = \frac{A_{k+1}- A_k}{A_{k}}$. Define recursively,
\begin{subequations}\label{Eq:QuasiSub2}
\begin{align}
z_{k+1} &= \arg \min_{z \in \X} \left\{\langle \nabla f(x_k),z \rangle + \frac{1}{\alpha_k} D_h(z, z_k)\right\}\\
x_{k+1} &= \frac{\tau_k}{\tau_k - 1}z_{k+1} + \frac{1}{\tau_k - 1}  x_k
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This algorithm has optimality conditions,
\begin{subequations}
\begin{align}
z_{k+1} &= x_{k+1} - \frac{A_{k}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqOther1}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k}) \label{Eq:XSeqOther1}.
\end{align}
\end{subequations}
\noindent Choosing $A_k = k(k+1)/2$ results in an $O(1/k)$ convergence rate. Note, structurally this algorithm is very similar to the conditional gradient method (see~\ref{App:OtherDiscret} for details). The second method does not comport to a straight-forward discretization technique and is given by the updates,
\begin{algorithm}[H]
\caption{Method 2}
{\bf Assumptions:} $f, h$ are convex and differentiable. $h$ is strongly-convex and $f$ has bounded gradients, $\sup_{x\in\X}\|\nabla f(x)\| = G <\infty$. 
Let $A_0 = 1$, $x_0 = z_0$ and $\tau_k = \frac{A_{k+1}- A_k}{A_{k}}$. Define recursively,
\begin{subequations}\label{Eq:QuasiSub3}
\begin{align}
x_{k+1}  &= \frac{\tau_k}{\tau_k + 1} z_k + \frac{1}{\tau_k+1}x_k,\\%\label{Eq:ZSeqQuasi}\\
z_{k+1} &= \arg \min_{z \in \X} \left\{\langle \nabla f(x_{k+1}),z\rangle + \frac{1}{\alpha_k} D_h(z, z_k)\right\}.
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This algorithm has optimality conditions,
\begin{subequations}\label{Eq:QuasiSub3}
\begin{align}
z_{k} &= x_{k+1} + \frac{A_{k}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqMethod2}\\%\label{Eq:ZSeqQuasi}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k+1}).\label{Eq:XSeqMethod2}
\end{align} 
\end{subequations}
Notice, \eqref{Eq:QuasiSub3} is very similar to the quasi-monotone subgradient method~\eqref{Eq:QuasiSub1}. In particular, under the same assumptions, it obtains a $O(1/\sqrt{k})$ convergence guarantee. Furthermore, like the quasi-monotone subgradient method, it can be extended to functions which have bounded subgradients instead of full gradients and an additional weighting term can be added.

%%%%%%%%%%%%%%%%%%%%%ACCELERATED GRAD DESCENT %%%%%%%%%%%%%%
\subsection{Accelerated Mirror Descent}
\label{Sec:AccGrad}
We present two variants of the accelerated mirror descent algorithm (the accelerated gradient descent algorithm in the setting where $h = \frac{1}{2}\|\cdot \|^2$). Neither of these correspond to a straight-forward discretization of~\eqref{Eq:E} -- they both entail introducing an additional sequence $\{y_k\}$ to obtain a better bound on the error. The first was the version was introduced by Michel Baes~\cite{Baes09}:
\begin{algorithm}[H]
\begin{subequations}\label{Eq:AcceleratedGrad2}
\caption{Accelerated Mirror Descent (Weakly Convex Setting)}
{\bf Assumptions:} $f, h$ are convex and differentiable. $h$ is $1$-strongly convex and $f$ has smooth gradients  $\|\nabla^{2} f\|\leq L$\\
Choose $A_0 = 1$, $M>0$, $\tilde A_{k+1} = L^{-1} A_{k+1}$,  $\tau_k = \frac{\tilde A_{k+1}- \tilde A_k}{\tilde A_{k+1}}:= \frac{\alpha_k}{\tilde A_{k+1}}$ and $x_0 = z_0 = y_0$. Define recursively, 
\begin{align}
z_k &= \arg \min_{z\in \X} \left\{ \sum_{i=1}^k \alpha_i \langle \nabla f(y_i), z\rangle + D_h(z, z_0)\right\}\label{Eq:ZSeqAcc2}\\
y_k &= \arg \min_{x \in \X} \left \{ \langle \nabla f(x_k), x\rangle + \frac{ L}{4M}\|y - x_k\|^2\right\}\\\
x_{k+1} &= \tau_k z_k + (1- \tau_k)  y_k\label{Eq:XSeqAcc2}
\end{align}
\end{subequations}
\end{algorithm}
\noindent This algorithm satisfies the following optimality conditions,

\begin{subequations}
\begin{align}
z_{k} &= y_k + \frac{\tilde A_{k+1}}{\tilde A_{k+1} -\tilde A_{k}} (x_{k+1} - y_k),  \label{Eq:ZSeqAcc3}\\%\label{Eq:ZSeqAcc2}\\
\nabla h(z_{k+1}) - \nabla h(z_{k}) &= -(\tilde A_{k+1} -  \tilde A_{k})  \nabla f(y_{k+1})  \label{Eq:XSeqAcc3}\\ %\label{Eq:XSeqAcc}\\
\quad M \|L^{-1} \nabla f(y_{k})\|_\ast^{2} &\leq L^{-1}\langle\nabla f(y_{k}), x_{k} - y_{k}\rangle \label{Eq:YSeqAcc3}
\end{align}
\end{subequations}
The advantage of this version of the accelerated mirror descent algorithm is that it can be extended to accelerate higher-order gradient methods (the details of which we give in Appendix~\ref{App:ProofAccGrad}). The second version, originally introduced by Nesterov~\cite{Nesterov05} entails computing the mirror descent update~\eqref{Eq:ZSeqAcc2} from the gradient at $x$:
\begin{equation*}
z_k = \arg \min_{z\in \X} \left\{ \sum_{i=1}^k \alpha_i \langle \nabla  f(x_i), z\rangle + D_h(z, z_0)\right\}
\end{equation*}
which results in an algorithm with the following optimality conditions, 
\begin{subequations}
\begin{align}
z_{k} &= y_k + \frac{\tilde A_{k+1}}{\tilde A_{k+1} -\tilde A_{k}} (x_{k+1} - y_k), \label{Eq:ZSeqAcc4}\\
\nabla h(z_{k+1}) - \nabla h(z_{k}) &= -(\tilde A_{k+1} -  \tilde A_{k})  \nabla f(x_{k+1})\label{Eq:XSeqAcc4}\\
\quad f(y_k) - f(x_k)  &\leq  -\frac{1}{2L} \| \nabla f(x_{k})\|_\ast^{2} \label{Eq:YSeqAcc4}
\end{align}
\end{subequation}
As mentioned above, to obtain a convergence guarantee we have replaced the current state $x_k$ by another sequence $y_k$ which simply needs to satisfy~\eqref{Eq:YSeqAcc3}. Furthermore, we have replaced our sequence $A_{k+1}$ by $\tilde A_{k+1}$ which depends on the Lipschitz parameter. We make the same adjustments to the Lyapunov function~\eqref{Eq:DiscLyapFunc1} which we sum up in the following theorem:
\begin{theorem} \label{Thm:AccGrad}
Using the following Lyapunov function
\begin{equation}\label{Eq:LyapAccGrad}
E_k = \tilde A_k(  f(y_k) -  f(x)) + D_h(x, z_k),
\end{equation}
we can show
  % which has the effect of scaling the gradient by $\epsilon$ as well. 
\[E_{k+1} - E_k \leq \varepsilon_k\]
where the error scales in the following way 
\[\varepsilon_k =  \left(\frac{1}{2}(\tilde A_{k+1} -\tilde A_{k})^{2}- \tilde A_{k+1}M\right)\|\nabla f(y_{k+1})\|^2 \]
for the accelerated gradient method~\eqref{Eq:AcceleratedGrad2}.
\end{theorem}
Choosing $x = x^\ast$ and $A_k$ such that $\tilde A_{k+1}^{-1}(\tilde A_{k+1} -\tilde A_{k})^{2} \leq 2M$ so that the error is non-positive, we obtain the following convergence rate,
\begin{equation}
 f(y_k) -  f(x^\ast) \leq \frac{\tilde A_0(f(x_0) - f(x^\ast)) + D_h(x^\ast, x_0)}{\tilde A_k}.
\end{equation}
Note, this condition requires that $A_k$ can grow at most quadratically which gives an $O(1/k^2)$ rate of convergence.
We give a proof of both of these methods in Appendix~\ref{App:ProofAccGrad}.

@@

We briefly point out that the algorithm resulting from applying something like the implicit-Euler method to both \eqref{Eq:Frank1}  and \eqref{Eq:Frank2}, 

@@

results in algorithm for which we can show $E_{k+1} - E_k \leq 0$ using~\eqref{Eq:FrankWolfeLyap}. However, like most implicit discretization techniques applied to the momentum dynamic~\eqref{Eq:ELBreg}, this does not lead to an algorithm with manageable subproblems.
We now turn our attention to analyzing trust region Newton-like method introduced by Nesterov. This method is given by the following update, 

>>>>

\section{The Dynamical View of Momentum Methods}

@@

\section{Rate-Generating Lyapunov Functions}
Lyapunov's method~\cite{Lyapunov} is based on the idea of constructing a positive definite quantity $\E: \X \rightarrow \R$ which decreases along the trajectories of the dynamical system $\dot X_t = v(X_t)$:
\begin{equation*}
\dot \E(X_t) =\langle \nabla \E(X_t), v(X_t)\rangle < 0.
\end{equation*}
The existence of such a \emph{Lyapunov function} guarantees that the dynamical system converges: if the function is positive yet strictly decreasing along all trajectories, then the dynamical system must eventually approach a region where $\E(X)$ is minimal.  If this region coincides with the stationary points of the dynamics, then all trajectories must converge to a stationary point.

@@

\subsection{The Quasi-Monotone Subgradient Method}
\label{Sec:QuasiMono}
The quasi-monotone subgradient method was developed by Nesterov~\cite{Nesterov15} as an alternative to the dual averaging method.  It also achieves a $O(1/\sqrt{k})$ lower bound, for the class of methods with bounded subgradients. The advantage of the quasi-monotone method however is that a convergence-rate guarantee can be shown for the entire sequence of iterates instead of the average (or minimum) iterate, as is the case with the dual averaging method in the non-Euclidean setting. {\em We begin by studying the quasi-monotone method in the situation where we have full gradients, which we assume to be bounded}. We also set the dual averaging term $\gamma_k \equiv 1$ for convenience and give an analysis of the full algorithm in Appendix~\ref{App:ProofQuasiGen}.  With these minor modifications, the algorithm can be understood as the explicit Euler method applied to~\eqref{Eq:Z} and the implicit Euler method applied to~\eqref{Eq:X} (where now, we take $e^{\beta_t} = A_{k+1}$):

@@

\subsection{Other Discretizations}
We give examples of two other algorithms that are the result of discretizing the momentum dynamic~\eqref{Eq:DynaBeta}, and analyze them using the discretized Lyapunov function~\eqref{Eq:DiscLyapFunc1}. The proofs of these results can be found in the Appendix~\ref{App:OtherDiscret}. The first method is the result of applying explicit Euler method to~\eqref{Eq:X} and the implicit Euler method to~\eqref{Eq:Z}:
\begin{algorithm}[H]
\caption{Method 1}
{\bf Assumptions:} $f$ is smooth and $\X$ is convex and compact. \\
Let $A_0 = 1$, $x_0 = z_0$, $\alpha_k = A_{k+1} - A_k$ and $\tau_k = \frac{A_{k+1}- A_k}{A_{k}}$. Define recursively,
\begin{subequations}\label{Eq:QuasiSub2}
\begin{align}
z_{k+1} &= \arg \min_{z \in \X} \left\{\langle \nabla f(x_k),z \rangle + \frac{1}{\alpha_k} D_h(z, z_k)\right\}\\
x_{k+1} &= \frac{\tau_k}{\tau_k - 1}z_{k+1} + \frac{1}{\tau_k - 1}  x_k
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This algorithm has optimality conditions,
\begin{subequations}
\begin{align}
z_{k+1} &= x_{k+1} - \frac{A_{k}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqOther1}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k}) \label{Eq:XSeqOther1}.
\end{align}
\end{subequations}
\noindent Choosing $A_k = k(k+1)/2$ results in an $O(1/k)$ convergence rate. Note, structurally this algorithm is very similar to the conditional gradient method (see~\ref{App:OtherDiscret} for details). The second method does not comport to a straight-forward discretization technique and is given by the updates,
\begin{algorithm}[H]
\caption{Method 2}
{\bf Assumptions:} $f, h$ are convex and differentiable. $h$ is strongly-convex and $f$ has bounded gradients, $\sup_{x\in\X}\|\nabla f(x)\| = G <\infty$. 
Let $A_0 = 1$, $x_0 = z_0$ and $\tau_k = \frac{A_{k+1}- A_k}{A_{k}}$. Define recursively,
\begin{subequations}\label{Eq:QuasiSub3}
\begin{align}
x_{k+1}  &= \frac{\tau_k}{\tau_k + 1} z_k + \frac{1}{\tau_k+1}x_k,\\%\label{Eq:ZSeqQuasi}\\
z_{k+1} &= \arg \min_{z \in \X} \left\{\langle \nabla f(x_{k+1}),z\rangle + \frac{1}{\alpha_k} D_h(z, z_k)\right\}.
\end{align} 
\end{subequations}
\end{algorithm}
\noindent This algorithm has optimality conditions,
\begin{subequations}\label{Eq:QuasiSub3}
\begin{align}
z_{k} &= x_{k+1} + \frac{A_{k}}{A_{k+1} -A_{k}} (x_{k+1} - x_k),\label{Eq:ZSeqMethod2}\\%\label{Eq:ZSeqQuasi}\\
\nabla h(z_{k+1}) -\nabla h(z_k) &= -(A_{k+1} - A_{k})\nabla f(x_{k+1}).\label{Eq:XSeqMethod2}
\end{align} 
\end{subequations}
Notice, \eqref{Eq:QuasiSub3} is very similar to the quasi-monotone subgradient method~\eqref{Eq:QuasiSub1}. In particular, under the same assumptions, it obtains a $O(1/\sqrt{k})$ convergence guarantee. Furthermore, like the quasi-monotone subgradient method, it can be extended to functions which have bounded subgradients instead of full gradients and an additional weighting term can be added.

%%%%%%%%%%%%%%%%%%%%%ACCELERATED GRAD DESCENT %%%%%%%%%%%%%%
\subsection{Accelerated Mirror Descent}
\label{Sec:AccGrad}
We present two variants of the accelerated mirror descent algorithm (the accelerated gradient descent algorithm in the setting where $h = \frac{1}{2}\|\cdot \|^2$). Neither of these correspond to a straightforward discretization of~\eqref{Eq:E}---they both entail introducing an additional sequence $\{y_k\}$ to obtain a better bound on the error. The first was the version was introduced by Michel Baes~\cite{Baes09}:
\begin{subequations}\label{Eq:AcceleratedGrad2}
\caption{Accelerated Mirror Descent (Weakly Convex Setting)}
{\bf Assumptions:} $f, h$ are convex and differentiable. $h$ is $1$-strongly convex and $f$ has smooth gradients  $\|\nabla^{2} f\|\leq L$\\
Choose $A_0 = 1$, $M>0$, $\tilde A_{k+1} = L^{-1} A_{k+1}$,  $\tau_k = \frac{\tilde A_{k+1}- \tilde A_k}{\tilde A_{k+1}}:= \frac{\alpha_k}{\tilde A_{k+1}}$ and $x_0 = z_0 = y_0$. Define recursively, 
z_k &= \arg \min_{z\in \X} \left\{ \sum_{i=1}^k \alpha_i \langle \nabla f(y_i), z\rangle + D_h(z, z_0)\right\}\label{Eq:ZSeqAcc2}\\
y_k &= \arg \min_{x \in \X} \left \{ \langle \nabla f(x_k), x\rangle + \frac{ L}{4M}\|y - x_k\|^2\right\}\\\
x_{k+1} &= \tau_k z_k + (1- \tau_k)  y_k\label{Eq:XSeqAcc2}
\noindent This algorithm satisfies the following optimality conditions:
z_{k} &= y_k + \frac{\tilde A_{k+1}}{\tilde A_{k+1} -\tilde A_{k}} (x_{k+1} - y_k),  \label{Eq:ZSeqAcc3}\\%\label{Eq:ZSeqAcc2}\\
\nabla h(z_{k+1}) - \nabla h(z_{k}) &= -(\tilde A_{k+1} -  \tilde A_{k})  \nabla f(y_{k+1})  \label{Eq:XSeqAcc3}\\ %\label{Eq:XSeqAcc}\\
\quad M \|L^{-1} \nabla f(y_{k})\|_\ast^{2} &\leq L^{-1}\langle\nabla f(y_{k}), x_{k} - y_{k}\rangle \label{Eq:YSeqAcc3}
The advantage of this version of the accelerated mirror descent algorithm is that it can be extended to accelerate higher-order gradient methods (the details of which we give in Appendix~\ref{App:ProofAccGrad}). The second version, originally introduced by Nesterov~\cite{Nesterov05}, entails computing the mirror descent update~\eqref{Eq:ZSeqAcc2} from the gradient at $x$:
z_k = \arg \min_{z\in \X} \left\{ \sum_{i=1}^k \alpha_i \langle \nabla  f(x_i), z\rangle + D_h(z, z_0)\right\}
which results in an algorithm with the following optimality conditions, 
z_{k} &= y_k + \frac{\tilde A_{k+1}}{\tilde A_{k+1} -\tilde A_{k}} (x_{k+1} - y_k), \label{Eq:ZSeqAcc4}\\
\nabla h(z_{k+1}) - \nabla h(z_{k}) &= -(\tilde A_{k+1} -  \tilde A_{k})  \nabla f(x_{k+1})\label{Eq:XSeqAcc4}\\
\quad f(y_k) - f(x_k)  &\leq  -\frac{1}{2L} \| \nabla f(x_{k})\|_\ast^{2} \label{Eq:YSeqAcc4}
\end{subequation}
As mentioned above, to obtain a convergence guarantee we have replaced the current state $x_k$ by another sequence $y_k$ which simply needs to satisfy~\eqref{Eq:YSeqAcc3}. Furthermore, we have replaced our sequence $A_{k+1}$ by $\tilde A_{k+1}$ which depends on the Lipschitz parameter. We make the same adjustments to the Lyapunov function~\eqref{Eq:DiscLyapFunc1} which we sum up in the following theorem:
\begin{theorem} \label{Thm:AccGrad}
Using the following Lyapunov function
\begin{equation}\label{Eq:LyapAccGrad}
E_k = \tilde A_k(  f(y_k) -  f(x)) + D_h(x, z_k),
  % which has the effect of scaling the gradient by $\epsilon$ as well. 
\[E_{k+1} - E_k \leq \varepsilon_k\]
\[\varepsilon_k =  \left(\frac{1}{2}(\tilde A_{k+1} -\tilde A_{k})^{2}- \tilde A_{k+1}M\right)\|\nabla f(y_{k+1})\|^2 \]
for the accelerated gradient method~\eqref{Eq:AcceleratedGrad2}.
Choosing $x = x^\ast$ and $A_k$ such that $\tilde A_{k+1}^{-1}(\tilde A_{k+1} -\tilde A_{k})^{2} \leq 2M$ so that the error is non-positive, we obtain the following convergence rate,
 f(y_k) -  f(x^\ast) \leq \frac{\tilde A_0(f(x_0) - f(x^\ast)) + D_h(x^\ast, x_0)}{\tilde A_k}.
Note, this condition requires that $A_k$ can grow at most quadratically which gives an $O(1/k^2)$ rate of convergence.
We give a proof of both of these methods in Appendix~\ref{App:ProofAccGrad}.

@@

We briefly point out that the algorithm resulting from applying something like the implicit Euler method to both \eqref{Eq:Frank1}  and \eqref{Eq:Frank2}, 

@@

results in algorithm for which we can show $E_{k+1} - E_k \leq 0$ using~\eqref{Eq:FrankWolfeLyap}. However, like most implicit discretization techniques applied to the momentum dynamics~\eqref{Eq:ELBreg}, this does not lead to an algorithm with manageable subproblems.
We now turn our attention to analyzing the trust region Newton-like method introduced by Nesterov. This method is given by the following update, 
